Simple repression draft

### 2020-09-14 doc/resubmission/section_02_means.tex
implement Robs 20200913 comments

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no 
transcription for many of these states. The models are probabilistic with each
state assigned some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key axes: what states to
consider and how to compute the probabilities of those states.

>>>>

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no 
transcription for many of these states. This discretization of a potentially
continuous number of promoter states (due to effects such as supercoiling of 
DNA~\cite{Chong2014, Sevier2016}) is analogous to how the Monod-Wyman-Changeux
model of allostery coarse-grains continuous molecule conformations into a finite
number of states~\cite{Martins2011}. The models are probabilistic with each
state assigned some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key axes: what states to
consider and how to compute the probabilities of those states.

### 2020-09-14 doc/resubmission/section_03_beyond_means.tex
implement Robs 20200913 comments

In this section, our objective is to explore the same models considered in the
previous section, but now with reference to the the question of how well they
describe the distribution of gene expression levels, with special reference to
the variance in these distributions. To that end, we repeat the same pattern as
in the previous section by examining the models one by one. In particular
we will focus on the Fano factor, defined as the variance/mean. This metric
serves as a powerful discriminatory tool from the null model that the
steady-state mRNA distribution must be Poisson, giving a Fano factor of one.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit modeling of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. 
% \mmnote{Haven't found model 5 formulated quite like this in the literature,
% usually burstiness is handled like model 4, but I should probably search a bit
% more.} 
The right column shows the Fano factor $\nu$ (variance/mean) for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
cannot produce the single-cell data observed in part (B) without introducing
additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
determined with single-molecule mRNA Fluorescence \textit{in situ}
Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
cross comparison with data presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

@@

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts
are produced as a Poisson process from a single promoter state. This is the
picture from Jones et.\ al.~\cite{Jones2014} that was used to interpret a
systematic study of gene expression noise over a series of promoters designed to
have different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescence measurements, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For \mrm{challenging} the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its
simplicity, with only two parameters, the initiation rate $r$ and degradation
rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
other words, the model is not excessively complex for the data at hand. But for
many interesting questions, for instance in the recent
work~\cite{Razo-Mejia2020}, knowledge of means and variances alone is
insufficient, and a description of the full distribution of molecular counts is
necessary. For this we need a (slightly) more complex model than model 1 that
would allow us to incorporate the non-Poissonian features of constitutive
promoters directly into a master equation formulation.

\mrm{I added the word challenging instead of disproving because of the 
following comment from reviewer 1. But maybe more needs to be discussed about 
this?}

\mrm{
    “disproving the universal noise curve from So et. al. [59]” (page 18). How was the universal noise curve disproved? Universal noise/mean relations, similar to those reported in ref \#59, were repeatedly observed, in both bacteria and eukaryotes (see, e.g., PMID: 24311680, PMID: 25858977). Moreover, later reports showed that the trend is observed even when accounting for gene dosage effects (PMID: 26965629). Thus, the statement above requires clarification.
}

@@

Although physically intuitive given our understanding of how a single transcript
is produced, we are forced to discard both of these models on the basis that the
Fano factor for both of them is strictly $< 1$ as shown in
Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
detailed derivations). This is in disagreement with the experimental data
from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
of these unregulated promoters --which only differ on the promoter sequence--
show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
\textit{any} model in which transcription proceeds through a multistep cycle
must necessarily have $\nu<1$. The intuitive argument compares the waiting time
distribution to traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an exponential
distribution. The former is a convolution of multiple exponentials, and
intuitively the waiting time distribution for a multistep process should be more
peaked with a smaller fractional width than a single exponential with the same
mean. A less disperse waiting time distribution means transcription initations
are more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
precise version of the intuitive arguments in this paragraph).

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning definite
physical meaning to the two states in this model.
\mrm{I like this paragraph's philosphy, but I don't know if it should be cut out 
for the sake of shrinking the paper further.}

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation possible between
transcription factor association rates and burst durations, but this assumption
could very well break down, so we will have to keep it in mind when we infer
repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
later~\cite{Jones2014}.
\mrm{Another great paragraph that could potentially be removed.}

>>>>

In this section, our objective is to explore the same models considered in the
previous section, but now with reference to the the question of how well they
describe the distribution of gene expression levels, with special reference to
the variance in these distributions. To that end, we repeat the same pattern as
in the previous section by examining the models one by one. In particular
we will focus on the Fano factor, defined as the variance/mean. This metric
serves as a powerful discriminatory tool to compare our different models to the
null model that the steady-state mRNA distribution must be Poisson, resulting a
Fano factor of one.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit treatment of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. 
% \mmnote{Haven't found model 5 formulated quite like this in the literature,
% usually burstiness is handled like model 4, but I should probably search a bit
% more.} 
The right column shows the Fano factor $\nu$ (variance/mean) for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
cannot produce the single-cell data observed in part (B) without introducing
additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
determined with single-molecule mRNA Fluorescence \textit{in situ}
Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
cross comparison with data presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

@@

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts
are produced as a Poisson process from a single promoter state. This is the
picture from Jones et.\ al.~\cite{Jones2014} that was used to interpret a
systematic study of gene expression noise over a series of promoters designed to
have different strengths, but no regulation. This model insists that the
``true'' steady-state mRNA distribution is Poisson, implying the Fano factor
$\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute measured
deviations from Fano = 1 to intensity variability in fluorescence measurements,
gene copy number variation, and copy number fluctuations of the transcription
machinery, e.g., RNAP itself. In this picture, all the corrections to Poisson
behavior are derived as additive corrections to the Fano factor. For challenging
the ``universal noise curve'' from So et.\ al.~\cite{So2011}, this picture was
excellent. It is appealing in its simplicity, with only two parameters, the
initiation rate $r$ and degradation rate $\gamma$. In other words, the model is
not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, attributing
all deviations from the model to extrinsic noise sources, limits the kinds of
predictions that can be done. To make progress then we need a (slightly) more
complex model than model 1 that would allow us to incorporate the non-Poissonian
features of constitutive promoters directly into a master equation formulation.

@@

Although physically intuitive given our understanding of how a single transcript
is produced, we are forced to discard both of these models on the basis that the
Fano factor for both of them is strictly $< 1$ as shown in
Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
detailed derivations). This is in disagreement with the experimental data
from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
of these unregulated promoters --which only differ in their promoter sequence--
show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
\textit{any} model in which transcription proceeds through a multistep cycle
must necessarily have $\nu<1$. The intuitive argument compares the waiting time
distribution to traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an exponential
distribution. The former is a convolution of multiple exponentials, and
intuitively the waiting time distribution for a multistep process should be more
peaked with a smaller fractional width than a single exponential with the same
mean. A less disperse waiting time distribution means transcription initations
are more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
precise version of the intuitive arguments in this paragraph).

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning definite
physical meaning to the two states in this model.

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation possible between
transcription factor association rates and burst durations, but this assumption
could very well break down, so we will have to keep it in mind when we infer
repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
later~\cite{Jones2014}.
